Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-22 doc/report/report.tex
report

%Main ideas, abstract and concrete, translation, parse- and abstract trees.
%Example of abstract.
%We say that this is a function that takes as argument and returns..
%Categories, valency built in (problems with ställa - på, under, vid). Parameters, fields.
%Example of a piece with Swedish, the reader gets familiar with how it looks.
%Also show how information 'disappears', become strings.
%What kind of linguistic analyse, what is the purpose to cover. Mention other uses,
%ideas of anaphores, montague? dependent types, ontologies, cool pictures by Krasse.
%
%The resources, why, how many, how to use, projects. functors.
%
%Still under big development, evolving, growing.
%For this project, only (mostly) important to model Swedish without abandoning the
%resources too much. Use of the big grammar for translation, why it might be hard with
%semantics and lexicon.
%Research about how to parse with gf by Krasimir, experiments with English.

\subsection{Swedish}
Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
for what will be written in the grammar part.
Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag är stor, I am big
Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\section{The grammar}
Although we can't expect to get a full coverage, we would like a tillfredsställande
Explain what's in the resources, constructions present in most languages.

\subsection{GF for Swedish}
What was in the resources
description of the existing grammar, the types, the possibilities
VP, NP.
valency built in (problems with ställa - på, under, vid).
Determiners, quants, predets, The difference from how pronouns are regarded. 
Variants, nonexist.
coverage of standard Swedish. Been working on a grammar fragment based on Talbanken,
like cool people like Montague. May also be interesting for other purposes.
Swedish shares with Nor and Dan. Extra module, covering things like sw is a
v2-lang, focusing parts of sentences.

@@

Formal subjects \cite{SAG-19} is often used in Swedish
\enumsentence{Det sitter en fågel på taket}
\emph{`There is a bird sitting on the roof'}
or, literally
\emph{`It sits a bird on the roof'}.\\
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{*Det äter en fågel på taket.}
unless their in passive form
\enumsentence{Det äts en fågel på taket.}
There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{*Det sitter den fågel på taket. \\
              **Det sitter fåglarna på taket.}
The distinction lays in the definitness of the noun itself, (not the whole noun phrase..)
it must be indefinite, even though determiners are allowed. Jfr
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter samtliga fåglar på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The grammar handles this by looking at the determiner used
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|
The function combines a verb phrase a determiner and a noun phrase in to a clause.
\verb|SimpleVP| is a normal VP, but with the restriction that the main verb must be
intransitiv, a \verb|V|, (and passives?). If the  determiner requires a noun in determinened
form, the clause is put to NONEXIST.

@@

%% Refl 
Could varandra be thought of as reflexive for 'oss'?
varandra and sin/sitt/sig. Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}

>>>>

%Functor
%Main ideas, abstract and concrete, translation, parse- and abstract trees.
%Example of abstract.
%We say that this is a function that takes as argument and returns..
%Categories, valency built in (problems with ställa - på, under, vid). Parameters, fields.
%Example of a piece with Swedish, the reader gets familiar with how it looks.
%Also show how information 'disappears', become strings.
%What kind of linguistic analyse, what is the purpose to cover. Mention other uses,
%ideas of anaphores, montague? dependent types, ontologies, cool pictures by Krasse.
%
%The resources, why, how many, how to use, projects. functors.
%
%Still under big development, evolving, growing.
%For this project, only (mostly) important to model Swedish without abandoning the
%resources too much. Use of the big grammar for translation, why it might be hard with
%semantics and lexicon.
%Research about how to parse with gf by Krasimir, experiments with English.

\subsection{Swedish}
Swedish is a North-Germanic language spoken,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also one
a official language in Finland and approximately 10 million people speak Swedish.
Charasterictic for Swedish is the fact that it is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects that are
being empathazied. \\

\enumsentence{
\begin{tabular}{lll}
  Johan & gick & sakta på gatan.\\
\emph{Johan} & \emph{walked} & \emph{slowely on the street.}\\
  Sakta & gick & Johan på gatan.\\
\emph{Slowly} & \emph{walked} & \emph{Johan on the street.}\\
  På gatan & gick & Johan sakta.\\
\emph{On the street} & \emph{walked}  & \emph{Johan slowly.}\\
\end{tabular}}

Inverted word order marks questions
\enumsentence{Gick Johan på gatan?\\
        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
\enumsentence{Jag såg Anna. Hon hälsade \textbf{inte}.\\
        \emph{I saw Anna. She did not say hello.}\\
              Jag såg att Anna \textbf{inte} hälsade.\\
        \emph{I saw that Anna did not say hello.}}

Some other features typical for Swedish is it's use of passive, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb
\enumsentence{Han jagade\textbf{s} av ett lejon.\\
        \emph{He was hunted by a lion.}}
Jfr.
\enumsentence{Ett lejon jagade honom.\\
        \emph{A lion was hunting him.}}

Constructions with 'det är/var' ('it is/was') is very common\cite{H&H309d}:
\enumsentence{Det var roligt att höra.\\
        \emph{I'm glad to hear that.}}
'Det' \emph{it} is also used as formal subject for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{Det står en älg på fältet.\\
        \emph{There is a mouse on the field.}}

The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog sig. & Han såg sitt barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog honom. & Han såg hans barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\section{The grammar}
Although we can't expect to get a full coverage, we would like a tillfredsställande
Explain what's in the resources, constructions present in most languages. 
Been working on a grammar fragment based on Talbanken,
like cool people like Montague. May also be interesting for other purposes.
Swedish shares with Nor and Dan. Extra module, covering things like sw is a
v2-lang, focusing parts of sentences.

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses.
Covers sentences like
han har inte ätit två gula äpplen idag
word order, agreement, tempus, adverbs, adjectivs...
and relatives
han såg katten han tyckte om
Also info in the Extra module, where more language specific constructions
can be found.
honom ville hon inte tänka på.

Before describing what has been added, will give a short introductions to the
the cats of resource grammar and the sw ipmlementation in particular.

Most important now is NP and VP.
Starting from N, get to CN what is it, what can we do now.
To turn in into a NP need to fix determination, by Det. Dets have .. fixed,
but allow variantion in.. 
What is the difference? Some words on differences for pronouns.
Example, use vartiants {}.
There are also Quants, which .... 
Once we have a NP, can use Predets. What they are.

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.

%Besides the categories NP,VP,AP there are also the import Det and Quant, Predet.

%What was in the resources
%description of the existing grammar, the types, the possibilities
%VP, NP.
%valency built in (problems with ställa - på, under, vid).
%Determiners, quants, predets, The difference from how pronouns are regarded. 
%Variants, nonexist.
%coverage of standard Swedish.

@@

Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish
\enumsentence{Det sitter en fågel på taket}
\emph{`There is a bird sitting on the roof'}
or, literally
\emph{`It sits a bird on the roof'}.\\
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{*Det äter en fågel på taket.}
unless their in passive form
\enumsentence{Det äts en fågel på taket.}
%det finns/saknas/fattas
There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{*Det sitter den fågel på taket. \\
              **Det sitter fåglarna på taket.}
The distinction lays in the definitness of the noun itself, (not the whole noun phrase..)
it must be indefinite, even though determiners are allowed. Jfr
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The grammar handles this by looking at the determiner used
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|
The function combines a verb phrase a determiner and a noun phrase in to a clause.
\verb|SimpleVP| is a normal VP, but with the restriction that the main verb must be
intransitiv, a \verb|V|, (and passives?). If the  determiner requires a noun in determinened
form, the clause is put to NONEXIST.

@@

%% Refl 
Could varandra be thought of as reflexive for 'oss'?
varandra and sin/sitt/sig. Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.
